---
title: "Fat_18_samples_mbr"
author: "Erik"
date: "6 Februar 2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This analysis is for adipose tissue from the MIDY biobank. The data was 
generated using all 18 samples (742 samples were removed) in Maxquant 1.6.3.4 
with default setting and match between runs were ON.

# 1 Libraries

```{r}
library(tidyverse)
```

# 2 Data processing

## 2.1 
```{r func_1, echo = FALSE}
make_tibble_for_DEP <- function(file_name = "proteinGroups.txt") {
  # Library 
  library(tidyverse)
  
  # Read file
  data_raw <- read_tsv(file_name, col_types = list("Reverse" = col_character()))
  
  # Extract LFQ column names
  lfq_col_names <- str_extract(colnames(data_raw), "LFQ.+")
  lfq_col_names <- lfq_col_names[which(!is.na(lfq_col_names))]
  
  data_raw <- data_raw %>%
    dplyr::select(`Protein IDs`, `Majority protein IDs`,
                  `Fasta headers`, Peptides, 
                  `Razor + unique peptides`, `Unique peptides`, 
                  lfq_col_names,
                  `Only identified by site`, `Reverse`,
                  `Potential contaminant`)
  
  # Return object
  return(data_raw)
}
```

Read proteinGroups.txt output file from MaxQuant.

```{r read_data}
data_raw <- make_tibble_for_DEP(file_name = "proteinGroups_fat_match_between_runs_no742.txt")
glimpse(data_raw)
```

Remove reverse hits from data.

```{r func_2, echo = FALSE}
remove_reverse_hits <- function(tb) {
  # Library
  library(tidyverse)
  
  data_raw <- tb %>% 
    filter(is.na(Reverse))
  
  return(data_raw)
}
```

```{r rem_rev}
data_forward <- remove_reverse_hits(data_raw)
```

`r length(as_vector(data_raw[, 2])) - length(as_vector(data_forward[, 2]))` protein groups were removed.

## 2.2 Get gene names from bioDBnet

Create a vector with 5 IDs for every protein group, that can be used as an input
in bioDBnet[https://biodbnet-abcc.ncifcrf.gov/db/db2db.php]

```{r func3, echo = FALSE}
extract_five_protein_ids_for_biodb <- function(tb) {
  
  # Libraries
  library(dplyr)
  library(tidyr)
  
  
  protein_ids <- tb %>% 
    dplyr::select(`Majority protein IDs`) %>% 
    mutate(Protein_IDs = `Majority protein IDs`,
           protein_ID = str_split(Protein_IDs, ";")) %>% # Split protein group 
    dplyr::select(`Majority protein IDs`, protein_ID) %>% 
    unnest(protein_ID) %>% 
    group_by(`Majority protein IDs`) %>% 
    top_n(5) # Extracts IDs for first five ocurrences per protein group
  
  # Return tibble
  return(protein_ids)
}
```

```{r ids_for_genes}
ids_for_gene_names <- extract_five_protein_ids_for_biodb(data_forward)

write_tsv(ids_for_gene_names, "fat_18samples_matchbr_ids_for_gene_names.txt")
```

The refseq IDs were entered into bioDBnet to get gene names, and the gene names
were then imported from text into excel and then copied to the `fat_18samples_matchbr_ids_for_gene_names.txt` file.

Read `fat_18samples_matchbr_ids_for_gene_names.txt` and reformat it.


```{r func4, echo = FALSE}
spread_gene_names <- function(tb, value_col = "Gene Symbol") {
  # Libraries
  library(dplyr)
  library(tidyr)
  
  gene_names <- tb %>% 
    mutate(`Gene Symbol` = ifelse(`Gene Symbol` == "-", NA, `Gene Symbol`)) %>% # Replaces - with NAs
    group_by(`Majority protein IDs`) %>% 
    mutate(index = 1:n()) %>%  # Creates an index for use with spread
    select(-protein_ID) %>% 
    spread(key = "index", value = value_col) %>% 
    rename(gene_1 = `1`, gene_2 = `2`, gene_3 = `3`, gene_4 = `4`, gene_5 = `5`)
  
  # Return object
  return(gene_names)
}
```

```{r spread_names}
# Read file
gene_name_list <- read_tsv("fat_18samples_matchbr_gene_names_biodb.txt")

# Reformat gene name list
gene_names <- spread_gene_names(tb = gene_name_list)
```
Calculate how many genes have a locus in the first column and a gene name in the
second or third column.

```{r}
# Gene name in second column
gene_names %>% 
  filter(str_detect(gene_1, "LOC.+") & !str_detect(gene_2, "LOC.+"))

# Gene name in third column
gene_names %>% 
  filter(str_detect(gene_1, "LOC.+") & !str_detect(gene_3, "LOC.+"))

# Fourth column 
gene_names %>% 
  filter(str_detect(gene_1, "LOC.+") & !str_detect(gene_4, "LOC.+"))

# Fifth column
gene_names %>% 
  filter(str_detect(gene_1, "LOC.+") & !str_detect(gene_5, "LOC.+"))

# Check if any columns with NA have gene names
gene_names %>% 
  filter(is.na(gene_1))
```

There's a total of 9 rows with a gene name in a latter column but not the first.
There's also some contaminants with possible gene names and locus.

```{r func5, echo = FALSE}
replace_LOC_and_NA_with_gene_name <- function(tb) {
  # Libraries
  library(dplyr)
  
  gene_names <- tb %>% 
    mutate(gene_name = if_else(str_detect(gene_1, "LOC.+") & !str_detect(gene_2,"LOC.+"), gene_2, gene_1), # Replace LOC with 2nd name
           gene_name = if_else(str_detect(gene_name, "LOC.+") & !str_detect(gene_3,"LOC.+"), gene_3, gene_name), # 3rd
           gene_name = if_else(str_detect(gene_name, "LOC.+") & !str_detect(gene_4,"LOC.+"), gene_4, gene_name), # 4th
           gene_name = if_else(str_detect(gene_name, "LOC.+") & !str_detect(gene_3,"LOC.+"), gene_5, gene_name), # 5th
           # Replace NAs with gene name or locus (for contaminants)
           gene_name = if_else(is.na(gene_name) & !is.na(gene_2), gene_2, gene_name), # Replace NA with 2nd name
           gene_name = if_else(is.na(gene_name) & !is.na(gene_3), gene_3, gene_name),
           gene_name = if_else(is.na(gene_name) & !is.na(gene_4), gene_4, gene_name),
           gene_name = if_else(is.na(gene_name) & !is.na(gene_5), gene_5, gene_name),
           gene_name = replace_na(gene_name, gene_1)) %>% # Replaces artifact NAs with 1st name
    dplyr::select(`Majority protein IDs`, gene_name, everything())
  
  return(gene_names)
}
```

```{r}
# Replace locus and NA
gene_names <- replace_LOC_and_NA_with_gene_name(gene_names)

# write tsv
write_tsv(gene_names, "fat_18samples_matchbr_gene_names_for_GO.txt")
```

The gene names were entered into bioDBnet to extract gene ontology and KEGG 
pathways for each protein using their human annotations.

```{r}
# Read gene ontology data
ontologies_untidy <- read_tsv("fat_18samples_matchbr_GOs.txt")
```

Tidy and clean gene ontology columns.

```{r func_6, echo = FALSE}
tidy_gene_ontology_output_from_biodb <- function(tb) {
  
  # Libraries
  library(dplyr)
  library(stringr)
  library(tidyr)

  
  # Tidy data
  ontologies <- tb %>% 
    rename(bp = `GO - Biological Process`,
           cc =`GO - Cellular Component`,
           mf = `GO - Molecular Function`,
           kegg = `KEGG Pathway Info`) %>%
    dplyr::select(`Majority protein IDs`, bp, cc, mf, kegg) %>% 
    gather(bp, cc, mf, kegg, key = "type", value = "annotation") %>% # gathers all annotations into one column
    group_by(`Majority protein IDs`, type) %>% 
    nest() %>% 
    mutate(data = str_split(data, ";")) %>% # Separates all annotations per gene
    unnest() %>% 
    separate(data, into = c("annotation_id", "annotation"), sep = "[0-9] \\[") %>% 
    ungroup() %>%  
    group_by(`Majority protein IDs`, type) %>% 
    distinct(annotation_id, .keep_all = TRUE) %>% # Removes duplicated annotations per gene that for some reason gets created
    dplyr::filter(!is.na(`Majority protein IDs`)) # Removes NA artifacts in gene_name
  
  # Clean up strings in annotaion_id column (identifiers)
  ontologies$annotation_id <- str_replace_all(ontologies$annotation_id, 'list\\(annotation = \\"', "")
  ontologies$annotation_id <- str_replace_all(ontologies$annotation_id, 'list\\(annotation = c\\(\\"', "")
  ontologies$annotation_id <- str_replace_all(ontologies$annotation_id, '.*\\"\\)', "")
  
  # Clean up strings in annotation column (descriptions)
  ontologies$annotation <- str_replace_all(ontologies$annotation, "Name: ", "")
  ontologies$annotation <- str_replace_all(ontologies$annotation, "\\].*", "")
  ontologies$annotation <- str_replace_all(ontologies$annotation, "Pathway Title:", "")
  
  # Replace empty strings and ")" with NA in annotation_id column
  ontologies <- ontologies %>% 
    mutate(annotation_id = ifelse(annotation_id == "", NA, annotation_id),
           annotation_id = ifelse(annotation_id == ")", NA, annotation_id)) 
  
  # Return object
  return(ontologies)
}
```

```{r clean_ont}
# Clean ontologies
ontologies_tidy <- tidy_gene_ontology_output_from_biodb(ontologies_untidy)
```

Join gene names and ontologies together.

```{r}
# Join
names_ontologies_tidy <- full_join(gene_names, ontologies_tidy, by = "Majority protein IDs") 
```



```{r}

```

Convert GOs to perseus format.

```{r func_7, echo = FALSE}
convert_GOs_to_perseus_format <- function(tb) {
  
  # Libraries
  library(dplyr)
  library(tidyr)
  
  GOs_clean_untidy <- tb %>% 
    gather(annotation_id, annotation, key = "var", value = "val") %>%
    unite(type_var, type, var, sep = "_") %>% # Unites GO category and IDs
    group_by(`Majority protein IDs`, type_var) %>% 
    mutate(val = paste(val, collapse = ";")) %>% # Paste all GO categories per protein group and category type
    distinct() %>% # Removes all duplicates
    group_by(`Majority protein IDs`) %>% 
    spread(key = "type_var", value = "val")
  
  return(GOs_clean_untidy)
}
```

```{r}
GOs_perseus <- convert_GOs_to_perseus_format(ontologies_tidy)
```

Join maxquant data with gene names and GOs.

```{r}
gene_names_untidy <- gene_names %>% 
  unite(alt_gene_names, gene_1:gene_5, sep = ";")

genes_GOs_untidy <- full_join(gene_names_untidy, GOs_perseus, by = "Majority protein IDs")

full_join(data_raw, genes_GOs_untidy, by = "Majority protein IDs") %>% View()
```

```{r}
join_genes_GOs_mqdata <- function(genes_data, GO_data, mq_data) {
  
  # Libraries
  library(tidyr)
  library(dplyr)
  
  # Unite all gene name columns wiht ";" as separator
  gene_names_untidy <- genes_data %>% 
    unite(alt_gene_names, gene_1:gene_5, sep = ";")
  
  # Join gene names with GOs ready in perseus format
  genes_GOs_untidy <- full_join(gene_names_untidy, GO_data, by = "Majority protein IDs")
  
  # Join gene names and GOs with maxquant data
  mq_data <- full_join(mq_data, genes_GOs_untidy, by = "Majority protein IDs")
  
  return(mq_data)
}
```

```{r}
join_genes_GOs_mqdata(gene_names, GOs_perseus, data_raw)
```

